Perturbing SU($n$)-symmetric interactions

### 2020-03-10 collective_sun/sun_perturbation.tex
fix old/incorrect section reference

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the
permutationally symmetric manifold $\M_0$ by the two-body perturbation
$\V_2$ in \eqref{eq:perturbations} is simply the restriction of $\V_2$
onto $\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQPS \sum_O \sum_{p<q} w_{Opq} O_{1,2}
  \EQPS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-spin
operator $O_{pq}$.  If the two-spin operator
$O=X\otimes Y+Y\otimes X$, then
\begin{align}
  H_2^{(1)} \EQPS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} + \col{Y}\,\col{X} - \col{XY} - \col{YX}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary permutationally symmetric
state $\ket\psi\in\M_0$ and expand (see Section \ref{sec:eigenstates})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \sum_O \sum_{p<q}
  \sp{\sum_k \p{h_{pk} w_{Okq} + h_{qk} w_{Opk}}
    - \tilde h_{pq} w_{Opq}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \tilde h_{pq} \equiv h_p + h_q - 2 h_{pq},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ (indexed by $\Delta$) is an eigenvector
of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h}_2 \c \v{\m w} = \Delta \v{\m w},
  \label{eq:V2_eig}
\end{align}
where
\begin{align}
  \check{\m h}_2
  \equiv \sum_{p<q} \ket{\set{p,q}} \sp{\sum_{k\notin\set{p,q}}
    \p{h_{pk} \bra{\set{k,q}} + h_{qk} \bra{\set{p,k}}}
    - \tilde h_{pq} \bra{\set{p,q}}}.
  \label{eq:V2_eig_mat}
\end{align}
We explicitly exclude $k\in\set{p,q}$ in \eqref{eq:V2_eig_mat} because
these terms vanish in \eqref{eq:H_V2_psi} from the fact that
$h_{kk}=w_{Okk}=0$.  Solving the general eigenvalue problem in
\eqref{eq:V2_eig} requires diagonalizing the
${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$ matrix
$\check{\m h}_2$.  When both he interaction Hamiltonian $H_0$
translationally invariant, however, this eigenvalue problem can be
reduced to that of diagonalizing a matrix of dimensions
$\sim N\times N$; we perform this reduction in Section
\ref{sec:exploiting_symmetries}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Translational invariance}

>>>>

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the
permutationally symmetric manifold $\M_0$ by the two-body perturbation
$\V_2$ in \eqref{eq:perturbations} is simply the restriction of $\V_2$
onto $\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQPS \sum_O \sum_{p<q} w_{Opq} O_{1,2}
  \EQPS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-spin
operator $O_{pq}$.  If the two-spin operator
$O=X\otimes Y+Y\otimes X$, then
\begin{align}
  H_2^{(1)} \EQPS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} + \col{Y}\,\col{X} - \col{XY} - \col{YX}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary permutationally symmetric
state $\ket\psi\in\M_0$ and expand (see Section \ref{sec:eigenstates})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \sum_O \sum_{p<q}
  \sp{\sum_k \p{h_{pk} w_{Okq} + h_{qk} w_{Opk}}
    - \tilde h_{pq} w_{Opq}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \tilde h_{pq} \equiv h_p + h_q - 2 h_{pq},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ (indexed by $\Delta$) is an eigenvector
of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h}_2 \c \v{\m w} = \Delta \v{\m w},
  \label{eq:V2_eig}
\end{align}
where
\begin{align}
  \check{\m h}_2
  \equiv \sum_{p<q} \ket{\set{p,q}} \sp{\sum_{k\notin\set{p,q}}
    \p{h_{pk} \bra{\set{k,q}} + h_{qk} \bra{\set{p,k}}}
    - \tilde h_{pq} \bra{\set{p,q}}}.
  \label{eq:V2_eig_mat}
\end{align}
We explicitly exclude $k\in\set{p,q}$ in \eqref{eq:V2_eig_mat} because
these terms vanish in \eqref{eq:H_V2_psi} from the fact that
$h_{kk}=w_{Okk}=0$.  Solving the general eigenvalue problem in
\eqref{eq:V2_eig} requires diagonalizing the
${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$ matrix
$\check{\m h}_2$.  When the interaction Hamiltonian $H_0$ and
perturbation $\V_M$ are translationally invariant, however, this
eigenvalue problem can be reduced to that of diagonalizing a matrix of
dimensions $\sim N\times N$; we perform this reduction in Section
\ref{sec:trans_inv}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Translational invariance}
\label{sec:trans_inv}
